# Annotate Observations {#sec-annotate-observations}

`ggmark()` can be used to add annotation plot for the selected observations.
`ggmark` accepts `mark` argument, which should be a `mark_draw()` object to
define how to draw the links. 

Currently, two helper functions are provided to generate these links:

- `mark_line()`: draws a line to the annotation panel.
- `mark_tetragon()`: draws a quadrilateral shape connecting observations and the panel.

All of these functions specify links as `pair_links()` as introduced in @sec-links.
Each pair of links will introduce a panel in ggmark to annotate these
observations.

```{r setup}
#| code-fold: true
library(ggalign)
set.seed(123)
small_mat <- matrix(rnorm(56), nrow = 7)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## plot data
By default, if no observations are explicitly selected, `ggmark()` selects all
observations and splits them based on the layout's grouping.

Calling `ggmark()` initializes a ggplot object, the underlying data is created
using `fortify_data_frame()`. Please refer to it for more details. In addition,
the following columns will be added to the data frame:

 - `.panel`: the panel for the aligned axis. It means `x-axis` for vertical
   stack layout (including top and bottom annotation), `y-axis` for
   horizontal stack layout (including left and right annotation).
 - `.names` and `.index`: a character names (only
   applicable when names exists) and an integer of index of the original data.
 - `.hand`: A factor with levels `c("left", "right")` for horizontal stack
   layouts, or `c("top", "bottom")` for vertical stack layouts, indicating
   the position of the linked observations.

::: {.grid}

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_line()) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_tetragon()) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

:::

You can control the size or reach of the annotation links by modifying the plot
margins using the `theme(plot.margin)` argument.

## Selecting Observations
You can manually specify which observations to annotate. Note that the
observations are based on data rows (not columns).

> Note: Only data for the selected observations are retained

::: {.grid}

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_line(1:3, c(3, 5:7))) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_tetragon(1:3, c(3, 5:7))) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

:::

You can specify the names (rownames) of the observations:

::: {.grid}

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_line(paste0("row", 1:3), paste0("row", c(3, 5:7)))) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_tetragon(paste0("row", 1:3), paste0("row", c(3, 5:7)))) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

:::

## Add Layout Grouping Information
you can use the `group1` and `group2` arguments to control whether the layout
panel groups and their ordering should be added too.

::: {.grid}

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_line(1:3), group1 = TRUE) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_tetragon(1:3), group1 = TRUE) +
    geom_boxplot(aes(.names, value)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

:::


## Facet
By default, `ggmark()` uses `facet_wrap` to define the facet, and you can use it
to control the facet apearance (just ignore the `facets` argument). We prefer
`facet_wrap()` here because it offers flexibility in positioning the strip on
any side of the panel, and typically, we only want to a single dimension to
create the annotate the selected observations. However, you can still use
`facet_grid()` to create a two-dimensional plot. Note that for horizontal stack
layouts, the row facets, or for vertical stack layouts, the column facets will
always be overwritten.

```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(mark_line()) +
    geom_boxplot(aes(.names, value, fill = .names)) +
    facet_wrap(vars(), scales = "free", strip.position = "right") +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```

## Customizing Link Appearance
You can further customize the appearance of link lines and quadrilaterals using
the element argument in `mark_line()`/`mark_tetragon()` function:

 - Link lines can be customized using the `element_line()`.
 - Link ranges can be customized using the `element_polygon()`.

By default, vectorized fields in `element_line()` and `element_polygon()` will
be recycled to match the total number of groups.

::: {.grid}

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(
        mark_line(4:6, 1:2, .element =  element_line(color = c("red", "blue")))
    ) +
    geom_boxplot(aes(.names, value, fill = .names)) +
    facet_wrap(vars(), scales = "free", strip.position = "right") +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

::: {.g-col-6}
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(
        mark_tetragon(4:6, 1:2, 
            .element =  element_polygon(fill = c("red", "blue"), alpha = 0.5)
        )
    ) +
    geom_boxplot(aes(.names, value, fill = .names)) +
    facet_wrap(vars(), scales = "free", strip.position = "right") +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
::: 

::: 

You can wrap the `element` with `I()` to recycle it to match the drawing groups.
The drawing groups typically correspond to the number of observations for
`element_line()`, as each observation will be linked with the plot panel. 

```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(
        mark_line(4:6, 1:2, 
            .element =  I(element_line(color = c("red", "blue")))
        )
    ) +
    geom_boxplot(aes(.names, value, fill = .names)) +
    facet_wrap(vars(), scales = "free", strip.position = "right") +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```

For `element_polygon()`, the drawing groups usually align with the defined
groups. However, if the defined group of observations is separated and cannot be
linked with a single quadrilateral, the number of drawing groups will be larger
than the number of defined groups.
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(
        mark_tetragon(4:6, 1:2, 
            .element =  I(element_polygon(fill = c("red", "blue"), alpha = 0.5))
        )
    ) +
    geom_boxplot(aes(.names, value, fill = .names)) +
    facet_wrap(vars(), scales = "free", strip.position = "right") +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```

For `stack_discrete()`, we usually don’t need to specify both hands in the
formula, since they are expected to share the same ordering and group structure.
This is because all plots within `stack_discrete()` maintain a common ordering
index across the layout.

However, explicitly specifying another hands becomes useful in `stack_cross()`, where
different observation orderings are involved. This scenario is covered in
@stack-cross.

```{r}
stack_discreteh(small_mat) +
    align_dendro(aes(color = branch), k = 3L) +
    scale_x_reverse(expand = expansion()) +
    theme(plot.margin = margin()) +
    ggmark(mark_line(4:6 ~ waiver(), 1:2 ~ waiver())) +
    geom_boxplot(aes(.names, value, fill = .names)) +
    theme(plot.margin = margin(l = 0.1, t = 0.1, r = 0.1, b = 0.1, unit = "npc")) +
    align_dendro(aes(color = branch), k = 3L) +
    scale_x_continuous(expand = expansion()) +
    theme(plot.margin = margin())
```

## mark_draw()
Both `mark_line()` and `mark_tetragon()` are built on top of `mark_draw()`
(strictly, `.mark_draw()`). This function allows you to define custom mark
styles by supplying a drawing function which must return a `grob`/`gList`.

The function passed to `mark_draw()` must accept two arguments:

- A data frame representing the panel-side coordinates (ggmark plot)

- A data frame representing the observation-side coordinates

Each observation is assumed to occupy a unit length in the layout. Therefore,
the observation-side coordinates include two terminal points `x`, `y`, `xend`,
`yend`—representing the start and end of the observation along the linking axis.

Additional columns in the `link` data frame include:

- `link_id`: The identifier of the link (e.g., following example `4:6` link has id "a").

- `link_panel`: Indicates which panel the link is drawn to, based on the layout.

- `link_index`: The layout index for positioning.

- `.hand`: Either "left"/"right" (horizontal) or "top"/"bottom" (vertical), specifying the hand of the observation.

- `.index`: The original index of the observation.

Here is an example that prints the structure of the panel and link data frames:

```{r}
set.seed(123)
p <- ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(
        mark_draw(function(panel, link) {
            print(panel)
            print(link)
        }, a = 4:6, 1:2)
    )
pdf(NULL)
print(p)
invisible(dev.off())
```

Here, we draw a triangle to connect the 
```{r}
set.seed(123)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(hjust = 0, angle = -60)) +
    anno_right() +
    align_kmeans(3L) +
    ggmark(
        mark_draw(function(panel, link) {
            x <- c((panel$x + panel$xend) / 2L, link$x, link$xend)
            y <- c((panel$y + panel$yend) / 2L, link$y, link$yend)
            grid::polygonGrob(x, y)
        }, 4, 2) # selecting one observation from each group for simple example
    ) +
    geom_boxplot(aes(.names, value, fill = .names)) +
    facet_wrap(vars(), scales = "free", strip.position = "right") +
    theme(plot.margin = margin(l = 0.1, t = 0.1, unit = "npc"))
```
