# Fine-Grained Alignment Control {#sec-fine-align-control}

The `ggalign` package provides six `free_` functions that give you precise
control over plot alignment behavior by selectively disabling alignment
constraints for specific plot elements in `align_plots()`. 

- `free_align()` - Disables panel alignment for specified axes.
- `free_space()` - Removes space constraints when aligning.
- `free_border()` - Attaches border elements (like axis titles) to the panel but
    keeps their size. 
- `free_lab()` - Similar to `free_border()`, but only applies to axis labels and
    titles.
- `free_vp()` - Customizes viewport placement.
- `free_guide()` - Overrides `guides` collection behavior

```{r setup}
#| code-fold: true
library(ggalign)
```

## Position String Notation

Most `free_` functions use a consistent `"tlbr"` notation system where
each letter represents a side of the plot:

- `t`: top
- `l`: left  
- `b`: bottom
- `r`: right

First, let's create some example plots:
```{r}
# A simple scatter plot
p_simple <- ggplot(mtcars) +
    geom_point(aes(mpg, disp))

# Create a plot with long axis labels that causes alignment issues
p_wide_labels <- ggplot(mtcars) +
    geom_bar(aes(y = factor(gear), fill = factor(gear))) +
    scale_y_discrete(
        name = NULL,
        labels = c(
            "3 gears are often enough",
            "But, you know, 4 is a nice number",
            "I would def go with 5 gears in a modern car"
        )
    )
```

## `free_align()`

By default, `align_plots()` aligns plot panels so axes match. This can cause
excessive spacing when plots differ in axis label lengths. `free_align()` disables
panel alignment on specified sides.

```{r}
# Problem: When combined, it looks bad due to forced alignment
align_plots(p_simple, p_wide_labels, ncol = 1L)
```

```{r}
# Solution: Use free_align to prevent panel alignment
align_plots(p_simple, free_align(p_wide_labels), ncol = 1L)
```

```{r}
# Partial solution: Free only the left side to maintain right alignment
align_plots(p_simple, free_align(p_wide_labels, axes = "l"), ncol = 1L)
```

## `free_space()` - Space Constraint Removal

`free_space()` removes the allocated space for plot elements when aligning. This
is particularly useful for removing padding introduced by long axis labels or
titles.

```{r}
align_plots(p_simple, free_space(p_wide_labels, spaces = "l"), ncol = 1L)
```

While it may not appear impactful on its own, `free_space()` is often essential
when used together with `free_border()`.

## `free_border()`
`free_border()` attaches borders (e.g., axis titles, tick marks) directly to the
plot panel. This keeps them visually close to the panel during alignment.
```{r}
align_plots(free_border(p_simple, borders = "l"), p_wide_labels, ncol = 1L)
```

However, `free_border()` does not remove the space these components occupy. This
means that, although the border (e.g., axis title) follows the plot panel more
tightly in `p_simple`, the space reserved for it still exists. During alignment,
this space is added to `p_wide_labels`, potentially adding unnecessary spaces.

To fully eliminate the size contribution of the border, we often combine
`free_border()` with `free_space()`:

```{r}
align_plots(
    free_space(free_border(p_simple, borders = "l"), spaces = "l"),
    p_wide_labels,
    ncol = 1L
)
```

This removes the reserved size on the left side (spaces = "l"), ensuring clean,
panel-based alignment without extra padding.

## `free_lab()`

`free_lab()` is similar to `free_border()`, but only attaches axis titles and tick
labels, not full borders. It’s mainly included for completeness; in most cases,
combining `free_border()` and `free_space()` is sufficient.

## `free_vp()`

The `free_vp()` function allows you to customize the grid viewport when aligning
plots, giving you precise control over plot positioning and sizing. This is
useful for precisely placing plots when alignment alone is insufficient.

```{r}
p1 <- ggplot(mtcars) +
    geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) +
    geom_boxplot(aes(gear, disp, group = gear))
```

::: {.grid}

::: {.g-col-6}
```{r}
# Blank line at the beginning to keep height consistent


align_plots(p1, p2, ncol = 2L)
```
:::

::: {.g-col-6}
```{r}
align_plots(p1, 
    free_vp(p2, height = unit(0.8, "npc"), just = "bottom", y = 0), 
    ncol = 2L
)
```
::: 

:::

## `free_guide()`

The `free_guide()` function allows you to override the `guides` argument for a
single plot.
```{r}
p_right <- ggplot(mtcars) +
    geom_point(aes(hp, wt, colour = mpg)) +
    patch_titles("right") +
    labs(color = "right")
p_top <- p_right +
    patch_titles("top") +
    scale_color_continuous(
        name = "top",
        guide = guide_colorbar(position = "top")
    )
p_left <- p_right +
    patch_titles("left") +
    scale_color_continuous(
        name = "left",
        guide = guide_colorbar(position = "left")
    )
p_bottom <- p_right +
    patch_titles("bottom") +
    scale_color_continuous(
        name = "bottom",
        guide = guide_colorbar(position = "bottom")
    )
align_plots(
    free_guide(p_right, NULL),
    free_guide(p_bottom, NULL),
    free_guide(p_top, NULL),
    free_guide(p_left, NULL),
    guides = "tlbr"
)
```

You can also specify which guide positions to be collected for individual plots.
```{r}
align_plots(
    free_guide(p_right, "r"),
    free_guide(p_bottom, "b"),
    free_guide(p_top, "t"),
    free_guide(p_left, "l")
)
```
