# Basic example

This chapter shows simple usage examples of `ggalign`.

```{r setup}
#| code-fold: true
library(ggalign)
```

## Data-Free Composition
The usage of data-free composition is simple: just create plots first then
arrange plots together!

```{r}
# Create individual plots
p1 <- ggplot(mtcars) +
    geom_point(aes(mpg, disp)) +
    ggtitle("Scatter Plot")

p2 <- ggplot(mtcars) +
    geom_boxplot(aes(factor(cyl), mpg)) +
    ggtitle("Box Plot")

p3 <- ggplot(mtcars) +
    geom_histogram(aes(mpg), bins = 15) +
    ggtitle("Histogram")

# Arrange plots in a grid
align_plots(p1, p2, p3, ncol = 2)
```

The `align_plots()` function offers flexible control over grid dimensions,
sizing, and layout specifications. You can control the arrangement with
parameters like `ncol`, `nrow`, `widths`, `heights`, and even use `area()`
specifications for complex layouts.

`ggalign` provides various options to control layout specifications, including
whether to collect legend guides, align axes, and more. For more, see [Part
`Data-Free Composition`].

## Data-Aware Composition

A common use case for data-aware composition is combining a heatmap with a
dendrogram. The dendrogram reveals hierarchical relationships among the data
(e.g., samples or genes), and the heatmap is reordered to align with the
dendrogram structure—ensuring consistent interpretation.

```{r}
set.seed(123)
# Prepare heatmap matrix
mat <- matrix(
    rnorm(200, mean = 5, sd = 2),
    nrow = 20, ncol = 10,
    dimnames = list(paste0("G", 1:20), paste0("S", 1:10))
)
```

Create a basic heatmap:
```{r}
ggheatmap(mat)
```

With `ggalign`, you can add elements using the same `+` syntax as in `ggplot2`.
For example, to add a dendrogram above the heatmap:
```{r}
ggheatmap(mat) +                                                # <1>
    anno_top() +                                                # <2>
    align_dendro()                                              # <3>
```

  1. We initialize a heatmap layout.
  2. we initialize an annotation in the top side of the heatmap body.
  3. Reorder and group the observations based on hierarchical clustering, and
  add a dendrogram tree.

This automatically reorders the heatmap rows or columns to reflect the
hierarchical structure in the dendrogram.

While data-aware composition is the core strength of `ggalign`, its full
capabilities go beyond a single example. For more advanced features and finer
control, see [Part `Data-Aware Composition`].
