# Get Started {#sec-get-started}

This chapter shows simple usage examples of `ggalign`.

```{r setup}
#| code-fold: true
library(ggalign)
```

## Data-Free Composition

The usage of data-free composition is simple: just create plots first then
arrange plots together!

```{r}
# Create individual plots
p1 <- ggplot(mtcars) +
    geom_point(aes(mpg, disp)) +
    ggtitle("Scatter Plot")

p2 <- ggplot(mtcars) +
    geom_boxplot(aes(factor(cyl), mpg)) +
    ggtitle("Box Plot")

p3 <- ggplot(mtcars) +
    geom_histogram(aes(mpg), bins = 15) +
    ggtitle("Histogram")

# Arrange plots in a grid
align_plots(p1, p2, p3, ncol = 2)
```

The `align_plots()` function offers flexible control over grid dimensions,
sizing, and layout specifications. You can control the arrangement with
parameters like `ncol`, `nrow`, `widths`, `heights`, and even use `area()`
specifications for complex layouts.

`ggalign` provides various options to control layout specifications, including
whether to collect legend guides, align axes, and more. For more, see
@sec-data-free.

## Data-Aware Composition

A common use case for data-aware composition is combining a heatmap with a
dendrogram. The dendrogram reveals hierarchical relationships among the data
(e.g., samples or genes), and the heatmap is reordered to align with the
dendrogram structure—ensuring consistent interpretation.

```{r}
set.seed(123)
# Prepare heatmap matrix
mat <- matrix(
    rnorm(200, mean = 5, sd = 2),
    nrow = 20, ncol = 10,
    dimnames = list(paste0("G", 1:20), paste0("S", 1:10))
)
```

Create a basic heatmap:
```{r}
ggheatmap(mat)
```

With `ggalign`, you can add elements using the same `+` syntax as in `ggplot2`.
For example, to add a dendrogram above the heatmap:
```{r}
ggheatmap(mat) +                                                # <1>
    anno_top() +                                                # <2>
    align_dendro()                                              # <3>
```

  1. We initialize a heatmap layout.
  2. we initialize an annotation in the top side of the heatmap body.
  3. Add a dendrogram tree in the top annotation, and Reorder and group the observations based on hierarchical clustering.

This automatically reorders the heatmap rows or columns to reflect the
hierarchical structure in the dendrogram.

While data-aware composition is the core strength of `ggalign`, its full
capabilities go beyond a single example. For more advanced features and finer
control, see @sec-data-aware.

## When to Use Each Approach

**Use data-free composition when:**
- Combining unrelated plots for publication figures
- Creating dashboard-style layouts
- Arranging plots with different data sources
- Simple spatial arrangement is sufficient

**Use data-aware composition when:**
- Analyzing the same dataset from multiple perspectives
- Creating heatmaps with annotations
- Ensuring observation consistency across plots
- Working with genomic, transcriptomic, or other omics data

This foundational understanding of ggalign's two composition modes will guide you through the more advanced features covered in subsequent chapters.


