# heatmap layout {#sec-heatmap-layout}
The `heatmap_layout()` function provides a powerful way to create customizable
heatmaps in R using `ggplot2`. This chapter will guide you through its usage.

`heatmap_layout()` is a specialized version of `quad_alignb()`, which itself is
a specific variant of `QuadLayout` (`quad_layout()`) designed to align
observations both horizontally and vertically. We introduce `heatmap_layout()`
directly, as it is more familiar to many users, especially those experienced
with popular heatmap packages like `pheatmap` and `ComplexHeatmap`.

```{r setup}
#| echo: false
#| output: false
library(ggalign)
set.seed(123)
small_mat <- matrix(rnorm(56), nrow = 7)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

```{r}
#| echo: false
ggside() +
    lapply(c("top", "left", "bottom", "right"), function(side) {
        c(
            list(quad_anno(side, initialize = TRUE)),
            lapply(seq_len(3L), function(i) {
                ggfree() +
                    ggtitle(sprintf("Plot %s", i)) +
                    theme(panel.background = element_rect(fill = scales::alpha("red", 0.5)))
            }),
            list(layout_annotation(theme = theme(
                plot.background = element_rect(color = "red", linewidth = unit(3, "cm"))
            )))
        )
    }) +
    quad_active() +
    ggtitle("Main Plot") +
    theme(panel.background = element_rect(fill = scales::alpha("red", 0.5))) +
    layout_title("QuadLayout") +
    layout_annotation(theme = theme(plot.title = element_text(size = 20)))
```

`heatmap_layout()` simplifies the creation of heatmap plots by integrating
essential elements for a standard heatmap layout, ensuring that the appropriate
data mapping and visualization layers are automatically applied. `ggheatmap()`
is an alias for `heatmap_layout()`.

```{r setup}
#| code-fold: true
#| eval: true
```

## input data
As mentioned in @sec-stack-layout-input-data, we typically require a matrix for
the `Layout` which need align observations. Internally, `fortify_matrix()` will
be used to process the data. You can provide a numeric or character vector, a
data frame, or any other data type that can be converted into a matrix using
`as.matrix()`.

```{r}
ggheatmap(small_mat)
```

## Main plot (heatmap body) {#sec-heatmap-layout-main-plot}
The `ggheatmap()`/`quad_layout()` functions arrange plots in the Quad-Side layout of
the main plot. When the layout is initialized, a `ggplot` object is automatically
created for the main plot.

For `ggheatmap()`, the matrix input will be converted into a long-format data
frame when drawing. The data in the underlying `ggplot` object includes the
following columns:

  - `.xpanel` and `.ypanel`: the column and row panel

  - `.x` and `.y`: the `x` and `y` coordinates

  - `.row_names` and `.column_names`: A factor of the row and column names of
    the original matrix (only applicable when names exist).

  - `.row_index` and `.column_index`: the row and column index of the original
    matrix.

  - `value`: the actual matrix value.

The default mapping will use `aes(.data$.x, .data$.y)`, but can be customized
using `mapping` argument.

By default, the main plot is regarded as the active plot, meaning you can add
ggplot2 elements directly to the main plot. 
```{r}
ggheatmap(small_mat) + 
    geom_point() + 
    scale_fill_viridis_c()
```

By default, `ggheatmap()` adds a heatmap layer. If the matrix has more than
20,000 cells (`nrow * ncol > 20000`), it uses `geom_raster()` for performance
efficiency; for smaller matrices, `geom_tile()` is used. You can explicitly
choose the layer by providing a single string (`"raster"` or `"tile"`) in the
`filling` argument.
```{r}
ggheatmap(small_mat, filling = "raster")
```

```{r}
ggheatmap(small_mat, filling = "tile")
```

Note, the filling layer will always use `mapping` of `aes(.data$.x, .data$.y)`,
if you want to customize filling, you can set `filling = NULL`, which will
remove the filling layer and allow you to add custom filling geoms.
```{r}
ggheatmap(small_mat, filling = NULL) +
    geom_tile(aes(fill = value), color = "black", width = 0.9, height = 0.9)
```

A heatmap pie charts can be easily drawn:
```{r}
set.seed(123)
ggheatmap(matrix(runif(360L), nrow = 20L), filling = NULL) +
    geom_pie(aes(angle = value * 360, fill = value))
```

For more complex customizations of pie charts, you can try using
`ggforce::geom_arc_bar()` instead.

## rasterization
When working with large heatmaps, it's often beneficial to rasterize the heatmap
body layer. You can achieve this by using the `raster_magick()` function. The
`res` argument controls the resolution of the raster image. By default, the
`res` argument matches the resolution of the current device, but specifying a
different value can help reduce the resolution of the rasterized heatmap body.
```{r}
ggheatmap(small_mat, filling = NULL) +
    raster_magick(geom_tile(aes(fill = value)), res = 50)
```

By leveraging `raster_magick()`, you can also perform image post-processing
using the `magick` package. This allows for custom image resizing with filters.
```{r}
ggheatmap(small_mat, filling = NULL) +
    # Use `magick::filter_types()` to check available `filter` arguments
    raster_magick(geom_raster(aes(fill = value)),
        magick = function(image) {
            magick::image_resize(image,
                # we resize to the 50% of width
                geometry = "50%x", filter = "Lanczos"
            )
        }
    )
```

Note: When using `magick::image_resize()`, you should specify the `geometry`
argument to resize the image. If only the `filter` is specified, it will only
distort the image data (though subtle). For more information on image resizing,
refer to
[ImageMagick's resize documentation](https://usage.imagemagick.org/resize/).

You can also rasterize all plots in the layout directly with `raster_magick()`.
This method is defined for both `ggheatmap()`/`quad_layout()` and
`stack_layout()` objects.

Additionally, You can use external packages like
[ggrastr](https://github.com/VPetukhov/ggrastr) or
[ggfx](https://ggfx.data-imaginist.com/reference/with_raster.html) to rasterize
the heatmap body.
```{r}
ggheatmap(small_mat, filling = FALSE) +
    ggrastr::rasterise(geom_tile(aes(fill = value)), dev = "ragg")
```

Likewise, you can also rasterize all plots in the layout directly with
`ggrastr::rasterise()` for both `ggheatmap()`/`quad_layout()` and
`stack_layout()`.
```{r}
ggrastr::rasterise(ggheatmap(small_mat), dev = "ragg")
```

Furthermore, [ggfx](https://ggfx.data-imaginist.com/) offers many image filters
that can be applied to ggplot2 layers. See the package for the details.

## annotations {#sec-heatmap-layout-annotations}
In `ggheatmap()`/`quad_layout()`, annotations are handled by a `stack_layout()`
object and can be positioned at the top, left, bottom, or right of the main plot
(heatmap body). 

By default, `ggheatmap()`/`quad_layout()` do not activate an annotation, You can
use `quad_anno()` to activate an annotation, directing all subsequent additions
to the specified annotation position. The `quad_anno()` function has the
following aliases:

  - `anno_top`: A special case of `quad_anno()` with `position = "top"`.
  - `anno_left`: A special case of `quad_anno()` with `position = "left"`.
  - `anno_bottom`: A special case of `quad_anno()` with `position = "bottom"`.
  - `anno_right`: A special case of `quad_anno()` with `position = "right"`.

When `quad_anno()` is added to a `ggheatmap()`/`quad_layout()`, it will try to
automatically create a new `stack_layout()`. For top and bottom annotations,
`stack_alignv()` or `stack_freev()` will be used; for left and right
annotations, `stack_alignh()` or `stack_freeh()` will be applied.

`quad_anno()` will always attempt to initialize a `stack_layout()` with the same
alignment as the current direction. This means that if observations need to be
aligned horizontally, `stack_alignh()` will be used for left and right
annotations, otherwise, `stack_freeh()` will be initialized instead. The same
logic applies for vertical alignment—`stack_alignv()` or `stack_freev()` will be
applied for top and bottom annotations, depending on whether alignment is
required. However, you can also manually add a `stack_free()` for directions
that require alignment, which I'll cover in the following section.

Additionally, `quad_anno()` will set the active context to the annotation. This
means that subsequent additions will be directed to the annotation rather than
the main plot. We use the term `active context` in contrast to `active plot` (as
described in @sec-stack-layout-layout-customize), since the annotation is a
`Layout` object.

```{r}
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    # we set the active context to the left annotation
    anno_left() +
    align_dendro()
```

By default, the annotation `stack_layout()` will try to inherit data from
`ggheatmap()`/`quad_layout()`. If the observations require alignment vertically,
this means the data from `ggheatmap()`/`quad_layout()` should be a matrix, the
column annotations will also require a matrix and the matrix from
`ggheatmap()`/`quad_layout()` will be transposed for use in the column annotations.
```{r}
ggheatmap(small_mat) +
    # we set the active context to the top annotation
    anno_top() +
    align_dendro()
```

You can further customize the layout design or add new plots in the annotation
stack, as described in @sec-stack-layout.
```{r}
ggheatmap(small_mat) +
    # in the heatmap body, we set the axis text theme
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    # we set the active context to the right annotation
    anno_right() +
    # in the right annotation, we add a dendrogram
    align_dendro(k = 3L) +
    # in the dendrogram, we add a point layer
    geom_point(aes(color = factor(branch)))
```

In this example:

 - `anno_right()` initialize the right annotation stack, and change the active
 context to the right of the heatmap.
 - `align_dendro(k = 3L)` adds a dendrogram to the annotation and sets itself as
 the active plot in the annotation stack. 
 - `geom_point(aes(color = factor(branch)))` is then added to this active plot
 within the annotation stack, here, it means the `align_dendro()` plot.


`ggheatmap()` aligns observations both horizontally and vertically, so it's safe
to always use `quad_anno()` directly, as all annotations require a matrix, and
the layout data is also a matrix. However, for `quad_alignh()` and
`quad_alignv()` (which I'll discuss in more detail in a @sec-quad-layout), which
only align observations in one direction, the data in the layout may not fit the
data for the annotation (when the layout requires alignment of observations, we
typically use a matrix, regardless of whether alignment is needed in one or two
directions)

    - `quad_alignh()`: aligning observations in horizontal direction, for column
    annotations, we ll need a data frame for `stack_free()`.
    - `quad_alignv()`: aligning observations in vertical direction, for row
    annotations, we ll need a data frame for `stack_free()`.

In both cases, `quad_anno()` won't initialize the annotation by default,
instead, you must provide the annotation `stack_layout()` manually.

## Adding stack layout {#sec-heatmap-layout-add-stack-layout}
Similar to adding a plot in `stack_layout()` (@sec-stack-layout), when the
direction requires alignment, you can add both `stack_align()` and
`stack_free()`. However, if the direction does not require alignment, you can
add only `stack_free()`.

To add a `stack_layout()` to the `ggheatmap()`, we must prevent the automatical
creation of annotation by `quad_anno()` by setting `initialize = FALSE`

::: {.grid}

::: {.g-col-6}
```{r}
my_stack_align <- stack_alignh(small_mat) +
    align_dendro(aes(color = branch), k = 3L)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(initialize = FALSE) +
    my_stack_align +
    layout_title("stack_align()")

```
:::

::: {.g-col-6}
```{r}
my_stack_free <- stack_freeh(mpg) +
    ggfree(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2)
ggheatmap(small_mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(initialize = FALSE) +
    my_stack_free +
    layout_title("stack_free()")
```
:::

:::

Note when aligning the observations, you must ensure the `number of observations`
is consistent in the direction. So for column annotations, you need transpose the data
manually.
```{r}
my_stack <- stack_alignv(t(small_mat)) +
    align_dendro(aes(color = branch), k = 3L)
ggheatmap(small_mat) +
    anno_top(initialize = FALSE) +
    my_stack
```

## `quad_active()`
To remove the active context and redirect additions back to the heatmap body,
you can use `quad_active()`. 
```{r}
ggheatmap(small_mat) +
    # we set the active context to the top annotation
    anno_top() +
    # we split the observations into 3 groups by hierarchical clustering
    align_dendro(k = 3L) +
    # remove any active annotation
    quad_active() +
    # set fill color scale for the heatmap body
    scale_fill_viridis_c()
```

## `quad_switch()`/`hmanno()`
We also provide `quad_switch()`/`hmanno()` (heatmap annotation) which integrates
`quad_active()` and `quad_anno()` into one function for ease of use. Feel free
to use any of these functions to streamline your annotation process.

::: {.grid}

::: {.g-col-6}
```{r}
ggheatmap(small_mat) +
    # we set the active context to the top annotation
    quad_switch("t") +
    # we split the observations into 3 groups by hierarchical clustering
    align_dendro(k = 3L) +
    # remove any active annotation
    quad_switch() +
    # set fill color scale for the heatmap body
    scale_fill_viridis_c() +
    layout_title("quad_switch()")
```
:::

::: {.g-col-6}
```{r}
ggheatmap(small_mat) +
    # we set the active context to the top annotation
    hmanno("t") +
    # we split the observations into 3 groups by hierarchical clustering
    align_dendro(k = 3L) +
    # remove any active annotation
    hmanno() +
    # set fill color scale for the heatmap body
    scale_fill_viridis_c()+
    layout_title("hmanno()")
```
:::

:::

## Plot Size
### Heatmap Body Size
You can specify the relative sizes of the heatmap body using the `width` and
`height` arguments in the `ggheatmap()` function.
```{r}
ggheatmap(small_mat, height = 2) +
    anno_top() +
    align_dendro()
```

Alternatively, the `quad_active()` function allows you to control the heatmap
body sizes. 
```{r}
ggheatmap(small_mat) +
    quad_active(height = 2) +
    anno_top() +
    align_dendro()
```

### Annotation Stack Size
The `quad_anno()` function allows you to control the total annotation stack
size. The `size` argument controls the relative width (for left and right
annotations) or height (for top and bottom annotations) of the whole annotation
stack.

```{r}
ggheatmap(small_mat) +
    anno_top(size = 1) +
    align_dendro()
```

You can also specify it as an absolute size using `unit()`:
```{r}
ggheatmap(small_mat) +
    anno_top(size = unit(30, "mm")) +
    align_dendro()
```

Note that the size of an individual plot ([@sec-stack-layout-plot-size]) does not
affect the total annotation stack size. You must adjust the annotation size using the
method described above.

::: {.grid}

::: {.g-col-6}
```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_dendro(size = unit(30, "mm")) +
    layout_title("plot size")
```
:::


::: {.g-col-6}

```{r}
ggheatmap(small_mat) +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    layout_title("annotation size")
```
:::

:::

In this chapter, we explored the usage of heatmap layout. These features
provide a strong foundation for visualizing matrix-based data in a structured
way. However, as your visualization needs grow more complex, the ability to
further customize and fine-tune the layout becomes essential.

In the next chapter, we will dive into the Layout Customize functionalities,
where you can gain full control over your plot's layout.
