# Layout customize
For layouts that can align observations, the package provides a suite of
`align_*` functions designed to give you precise control over the observations.
These functions allow you to reorder observations or partition them into
multiple groups.

Currently, there are four key `align_*` functions available for layout customization:

- **`align_group`**: Group and align plots based on categorical factors.
- **`align_order`**: Reorder layout observations based on statistical weights
 or allows for manual reordering based on user-defined ordering index.
- **`align_kmeans`**: Group observations by k-means clustering results.
- **`align_hclust`**: Reorder or group observations based on hierarchical clustering.

You can inspect the behavior of any `align_*` function by printing it. Note that
these functions do not add plot areas and set the active context, meaning you
cannot incorporate `ggplot2` elements directly into these objects.

```{r setup}
#| code-fold: true
#| collapse: true
library(ggalign)
set.seed(123)
small_mat <- matrix(rnorm(56), nrow = 7)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## `align_group()`
The `align_group()` function allows you to split the observations into groups.

```{r}
set.seed(1234)
ggheatmap(small_mat) +
    anno_top() +
    align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))
```

Note that all `align_*` functions which split observations into groups must not
break the previous established groups. This means the new groups must nest in
the old groups, usually they cannot be used if groups already exist.

```{r error=TRUE}
set.seed(1234)
ggheatmap(small_mat) +
    anno_top() +
    align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE)) +
    align_group(sample(letters[1:5], ncol(small_mat), replace = TRUE))
```

## `align_order()`
The `align_order()` function reorder the observations based on the summary
weights. 

In this example, we order the rows based on their means. By default, the
ordering is in ascending order according to the summary weights. You can reverse
the order by setting `reverse = TRUE`.

::: {.grid}

::: {.g-col-6}
```{r align_order}
ggheatmap(small_mat) +
    anno_left() +
    align_order(rowMeans) +
    layout_title(title = "reverse = FALSE")
```
:::

::: {.g-col-6}
```{r}
ggheatmap(small_mat) +
    anno_left() +
    align_order(rowMeans, reverse = TRUE) +
    layout_title(title = "reverse = TRUE")
```
:::

:::

Additionally, you can provide the ordering integer index or character index
directly:

```{r}
my_order <- sample(nrow(small_mat))
print(rownames(small_mat)[my_order])
```

::: {.grid}

::: {.g-col-6}
```{r}
ggheatmap(small_mat) +
    anno_left() +
    align_order(my_order)
```
:::

::: {.g-col-6}
```{r}
ggheatmap(small_mat) +
    anno_left() +
    align_order(rownames(small_mat)[my_order])
```
:::

:::

Some `align_*` functions also accept a `data` argument. It’s important to note
that all `align_*` functions treat rows as the observations. This means
`NROW(data)` must match the `number of observations` along the axis used for
alignment. The `data` argument can also accept a function (supporting purrr-like
lambda syntax), which will be applied to the layout matrix.

As mentioned in @sec-heatmap_layout, for top and bottom annotations, the data
matrix of `quad_layout()`/`ggheatmap()` is transposed to create the annotation
`stack_layout()`. Therefore, you can use `rowMeans()` to calculate the mean value
across all columns.

```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_order(rowMeans)
```

Some `align_*` functions that reorder observations include an argument called
`strict`. This argument is especially useful when previous groups have already
been established. If previous groups have been created and `strict = FALSE`, the
function will reorder the observations within each group.
```{r}
set.seed(1234)
ggheatmap(small_mat) +
    anno_top() +
    align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))+
    align_order(rowMeans, strict = FALSE)
```

## `align_kmeans()`
The `align_kmeans()` function split the observations into groups based on
k-means clustering.
```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_kmeans(3L)
```

## `align_hclust()`
The `align_hclust()` function is a specialized version of `align_dendro()`, designed
to reorder observations and group them based on hierarchical clustering. Unlike
`align_dendro()`, however, `align_hclust()` does not add a dendrogram tree to the
plot. All the arguments intruduced here can also be used by `align_dendro()`.

```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_hclust()
```

Hierarchical clustering is performed in two steps: calculate the distance matrix
and apply clustering. You can use the `distance` and `method` argument to
control the builind process.

There are two ways to specify `distance` metric for clustering:

  - specify `distance` as a pre-defined option. The valid values are the supported
  methods in `dist()` function and coorelation coefficient `"pearson"`,
  `"spearman"` and `"kendall"`. The correlation distance is defined as 
  `1 - cor(x, y, method = distance)`.
  - a self-defined function which calculates distance from a matrix. The function
  should only contain one argument.

```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_hclust(distance = "pearson") +
    layout_title("pre-defined distance method (1 - pearson)")
```

```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_hclust(distance = function(m) dist(m)) +
    layout_title("a function that calculates distance matrix")
```

Method to perform hierarchical clustering can be specified by `method`. Possible
methods are those supported in `hclust()` function. And you can also provide a
self-defined function, which accepts the distance object and return a `hclust`
object.

```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_hclust(method = "ward.D2")
```

You can specify `k` or `h` to split observations into groups, which work
similarly to `cutree()`: 

```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_hclust(k = 3L)
```

You can reorder the dendrogram based on the mean values of the observations by
setting `reorder_dendrogram = TRUE`.
```{r}
#| fig.width: 10
h1 <- ggheatmap(small_mat) +
    anno_top() +
    align_hclust(k = 3, reorder_dendrogram = TRUE) +
    layout_title("reorder_dendrogram = TRUE")
h2 <- ggheatmap(small_mat) +
    anno_top() +
    align_hclust(k = 3) +
    layout_title("reorder_dendrogram = FALSE")
align_plots(h1, h2)
```

In this example, we use `align_plots()` to arrange the plots, which functions
similarly to `cowplot::align_plots()` and `patchwork::wrap_plots()`. However,
unlike those, `align_plots()` can be directly used with quad_layout() and
stack_layout(), ensuring proper alignment by plot panel. Additionally,
`align_plots()` can align pheatmap and ComplexHeatmap objects, although they
won’t align by panel area in the same way as ggplot2 plots.

`align_hclust()` can also perform clustering between groups, meaning it can be
used even if there are existing groups present in the layout, in this way, you
cannot specify `k` or `h`: 

```{r}
set.seed(3L)
column_groups <- sample(letters[1:3], ncol(small_mat), replace = TRUE)
ggheatmap(small_mat) +
    anno_top() +
    align_group(column_groups) +
    align_hclust()
```

You can reorder the groups by setting `reorder_group = TRUE`.
```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_group(column_groups) +
    align_hclust(reorder_group = TRUE)
```

If you specify `k` or `h`, this will always turn off sub-clustering. The same
principle applies to `align_hclust()`, where new groups must be nested within
the previously established groups.
```{r error=TRUE}
ggheatmap(small_mat) +
    anno_top() +
    align_group(column_groups) +
    align_hclust(k = 2L)
```
