# ggoncoplot {#sec-ggoncoplot}

The `ggoncoplot()` function generates `oncoPrint` visualizations that display
genetic alterations in a matrix format. This function is especially useful for
visualizing complex genomic data, such as mutations, copy number variations, and
other genomic alterations in cancer research. 

```{r setup}
#| code-fold: true
library(ggalign)
```

## Input Data
The input should be a character matrix which encodes the alterations, you can use
string of `r ggalign:::oxford_and(c(";", ":", ",", "|"))` to separate multiple
alterations. Internally, `ggoncoplot()` will use `fortify_matrix()` to get such
matrix. 
```{r}
mat <- read.table(
    textConnection(
        "s1,s2,s3
         g1,snv;indel,snv,indel
         g2,,snv;indel,snv
         g3,snv,,indel;snv"
    ),
    row.names = 1, header = TRUE, sep = ",", stringsAsFactors = FALSE
)
mat
```

A basic oncoprint can be generated as follows:
```{r}
ggoncoplot(mat)
```

## OncoPrint Customization
By default, all alterations are represented with tiles of equal width and
height, which may lead to overlapping. You can control the `width` and `height`
of the tiles using the `map_width` and `map_height` arguments (we will introduce
another more effective ways to handle this in the `Specialized Geoms` section):
```{r}
ggoncoplot(mat, map_width = c(snv = 0.5), map_height = c(indel = 0.9))
```

By default, all empty string will be converted to `NA` value and `ggplot2` will
translate the `NA` values and render it in the legend. To prevent this, you can
turn off the translation using `na.translate = FALSE`: 
```{r}
ggoncoplot(mat, map_width = c(snv = 0.5), map_height = c(indel = 0.9)) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE)
```

The function automatically reorders rows and columns based on alteration
frequency. Disable this with `reorder_row = FALSE` and `reorder_column = FALSE`.

You can further customize layouts using methods from
@sec-layout-customize.
```{r}
ggoncoplot(mat,
    map_width = c(snv = 0.5), map_height = c(indel = 0.9),
    reorder_row = FALSE, reorder_column = FALSE
) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE)
```

## Advanced Data Handling
`ggoncoplot()` is built on `ggheatmap()` with a default `scheme_data()` (see
@sec-schemes for details), which splits alterations into separate
entries for visualization. (See `ggplot2 specification` section in `ggheatmap()`
for the input data of `scheme_data()`):
```{r eval=rlang::is_installed("tidyr")}
pdata <- function(data) {
    tidyr::separate_longer_delim(data,
        cols = value,
        delim = stringr::regex("\\s*[;:,|]\\s*")
    )
}
# Note: this figure will contain an empty string value.
#
# `ggoncoplot()` will automatically convert any empty strings to `NA`
# before pass it to `ggheatmap`.
ggheatmap(mat) -
    scheme_data(pdata) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE)
```

By default, the `scheme_data()` is inherited from the parent layout if the data of
the plot is inherited from the layout. You can apply the parent `scheme_data()`
first and then apply another transformation by setting `inherit = TRUE`. This
functionality is especially useful when working with `ggoncoplot()`.
```{r}
ggoncoplot(mat, map_width = c(snv = 0.5), map_height = c(indel = 0.9)) +
    anno_top(size = 0.2) +
    ggalign() +
    # by setting `inherti = TRUE`, we apply the parent layout `scheme_data()`
    # (from the `ggoncoplot()` layout) firstly, which will split the alteration
    # string and unnested the columns.
    # Here: We then remove `NA` value
    scheme_data(~ subset(.x, !is.na(value)), inherit = TRUE) +
    geom_bar(aes(.x, after_stat(count), fill = value)) +
    # note: `ggoncoplot()` use `geom_tile()` to draw the oncoPrint,
    # the guide is different from `geom_bar()`, though both looks
    # like the same, we remove the guide of `geom_bar()` plot
    guides(fill = "none") &
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE)
```

## Integration with `maftools`
The `ggoncoplot()` function supports `MAF` objects from the `maftools` package
using the `fortify_matrix.MAF()` method. It includes options to control data
generation for `ggoncoplot()`, such as drawing the top `n_top` genes.

```{r}
# load data from `maftools`
laml.maf <- system.file("extdata", "tcga_laml.maf.gz", package = "maftools")
# clinical information containing survival information and histology. This is optional
laml.clin <- system.file("extdata", "tcga_laml_annot.tsv", package = "maftools")
laml <- maftools::read.maf(
    maf = laml.maf,
    clinicalData = laml.clin,
    verbose = FALSE
)
# Visualizing the Top 20 Genes
ggoncoplot(laml, n_top = 20) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
    theme_no_axes("x")
```

By default, samples without alterations in the selected genes are removed. To
include all samples, set `remove_empty_samples = FALSE`: 
```{r}
align_plots(
    ggoncoplot(laml, n_top = 20L, remove_empty_samples = FALSE) +
        scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
        theme_no_axes("x") +
        ggtitle("Keep empty samples"),
    ggoncoplot(laml, n_top = 20L, remove_empty_samples = TRUE) +
        scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
        theme_no_axes("x") +
        ggtitle("Remove empty samples"),
    ncol = 1L,
    guides = "tlbr"
)
```

## Specialized Geoms
### `geom_subtile()`
When multiple alterations occur in the same sample and gene, they are combined
into a single value, `"Multi_Hit"`, by default. To visualize these alterations
separately, you can set `collapse_vars = FALSE`. However, doing so can lead to
overlapping alterations within the same cell, making the visualization cluttered
and hard to interpret.

In such cases, disabling the default filling and defining a custom heatmap layer
with `geom_subtile()` is more effective. This function subdivides each cell into
smaller rectangles, allowing the distinct alterations to be clearly displayed.
```{r}
ggoncoplot(laml, n_top = 20, collapse_vars = FALSE, filling = FALSE) +
    geom_subtile(aes(fill = value), ncol = 1L) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
    theme_no_axes("x")
```

We focus exclusively on cells with multiple alterations to highlight the
differences, by filtering the data before applying `geom_subtile()`:
```{r}
ggoncoplot(laml, n_top = 20, collapse_vars = FALSE, filling = FALSE) +
    scheme_data(function(data) {
        dplyr::filter(data, dplyr::n() > 1L, .by = c(".x", ".y"))
        # we apply the parent layout `scheme_data()` first (`inherit = TRUE`),
        # which will split the alteration string and unnested the columns.
    }, inherit = TRUE) +
    geom_subtile(aes(fill = value), ncol = 1L) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
    theme_no_axes("x")
```

### `geom_gshape()`
`geom_subtile()` often suffices for most scenarios. However, if you require a
strategy similar to that of `ComplexHeatmap`, consider using `geom_gshape()`,
which offers greater flexibility for complex customizations.

Key Features of `geom_gshape()`:

 - Custom Drawing Logic: Accepts a `gshape` aesthetic, allowing each cell to be
 rendered as a specific graphical object (`grob`). 
 - Flexible Aesthetic Mapping: You must provide the manual scaling
 `scale_gshape_manual()` to map `gshape` values to corresponding drawing
 functions. Functions mapped through `scale_gshape_manual()` can utilize any
 number of ggplot2 aesthetics and output custom graphical objects (`grob`).

The function provided in `values` argument of `scale_gshape_manual()` operates
similarly to `alter_fun` in ComplexHeatmap. Currently, `geom_subtile()` performs
well enough for most cases, making `geom_gshape()` somewhat cumbersome to use.
Managing alteration types, especially in the case of overlapping alterations,
requires meticulous design to ensure that the visual elements do not interfere
with each other.

## `ggalign` Attributes
Except for the data used for the main plot, `fortify_matrix.MAF()` also attaches
several useful attributes.

  - `gene_summary`: A data frame of gene summary informations. See
    `maftools::getGeneSummary()` for details.
  - `sample_summary`: A data frame of sample summary informations. See
    `maftools::getSampleSummary()` for details.
  - `sample_anno`: A data frame of sample clinical informations. See
    `maftools::getClinicalData()` for details.
  - `variant_weights`: A data frame of variant weights. Each gene in a sample
    is assigned a total weight of `1`. When multiple variants occur in the
    same gene-sample pair, the weight for each variant reflects its proportion
    of the total.
  - `n_genes`: Total number of genes.
  - `n_samples`: Total number of samples.
  - `titv`: A list of data frame with Transitions and Transversions
    summary. See `maftools::titv()` for details.

You can extract these attributes using the `ggalign_attr()` function. Below is
an example of how to extract the sample summary data (e.g., TMB) using
`ggalign_attr()` and add it to the top annotation of an oncoplot:
```{r}
ggoncoplot(laml, n_top = 20, collapse_vars = FALSE, filling = FALSE) +
    geom_subtile(aes(fill = value), ncol = 1) +
    theme_no_axes("x") +
    anno_top(size = 0.2) +
    ggalign(data = function(data) {
        data <- ggalign_attr(data, "sample_summary")
        # matrix input will be automatically melted into a long foramted data
        # frame in `ggalign()` function.
        as.matrix(data[2:(ncol(data) - 1L)])
    }) +
    geom_bar(aes(.x, value, fill = .column_names),
        stat = "identity"
    ) +
    ylab("TMB") &
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE)
```

We can draw the example from 
[maftools vignette](https://bioconductor.org/packages/release/bioc/vignettes/maftools/inst/doc/oncoplots.html).

```{r}
ggoncoplot(laml, n_top = 20, collapse_vars = FALSE, filling = FALSE) +
    geom_subtile(aes(fill = value), ncol = 1) +
    theme_no_axes("x") +
    # since legends from geom_tile (oncoPrint body) and `geom_bar`
    # is different, though both looks like the same, the internal
    # won't merge the legends. we remove the legends of oncoPrint body
    guides(fill = "none") +
    # add top annotation
    anno_top(size = 0.2) +
    ggalign(data = function(data) {
        data <- ggalign_attr(data, "sample_summary")
        # matrix input will be automatically melted into a long foramted data
        # frame in `ggalign()` function.
        as.matrix(data[2:(ncol(data) - 1L)])
    }) +
    geom_bar(aes(.x, value, fill = .column_names),
        stat = "identity"
    ) +
    ylab("TMB") +
    # add right annotation
    anno_right(size = 0.2) -
    # remove bottom spaces of the right annotation when aligning
    scheme_align(free_spaces = "b") +
    # add the text percent for the alterated samples in the right annotation
    ggalign(data = function(data) {
        # Atomic vector will be put in the `value` column of the data frame.
        ggalign_attr(data, "gene_summary")$AlteredSamples /
            ggalign_attr(data, "n_samples")
    }) +
    geom_text(aes(1, label = scales::label_percent()(value)), hjust = 1) +
    scale_x_continuous(
        expand = expansion(),
        name = NULL, breaks = NULL,
        limits = c(0, 1)
    ) +
    theme(plot.margin = margin()) +
    # add the bar plot in the right annotation
    ggalign(data = function(data) {
        data <- ggalign_attr(data, "gene_summary")
        # matrix input will be automatically melted into a long foramted data
        # frame in `ggalign()` function.
        as.matrix(data[2:8])
    }) +
    geom_bar(aes(value, fill = .column_names),
        stat = "identity",
        orientation = "y"
    ) +
    xlab("No. of samples") -
    # we apply the scale mapping to the top and right annotation: `position = "tr"`
    # and the main plot: `main = TRUE`
    with_quad(
        scale_fill_brewer("Mutations",
            palette = "Dark2", na.translate = FALSE
        ),
        position = "tr",
        main = TRUE
    ) +
    # add bottom annotation
    anno_bottom(size = 0.2) +
    # add bar plot in the bottom annotation
    ggalign(data = function(data) {
        data <- ggalign_attr(data, "titv")$fraction.contribution
        # matrix input will be automatically melted into a long foramted data
        # frame in `ggalign()` function.
        as.matrix(data[2:7])
    }) +
    geom_bar(aes(y = value, fill = .column_names), stat = "identity") +
    ylab("Ti/Tv") +
    scale_fill_brewer("Ti/Tv", palette = "Set2")
```

## Integration with GISTIC Results from `maftools`
The package also includes a `fortify_matrix.GISTIC()` method designed to handle
`GISTIC` objects from the `maftools` package. This allows you to seamlessly
apply the same operations to visualize `GISTIC` results. The following ggalign
attributes are generated as part of the analysis:

  - `sample_anno`: sample clinical informations provided in `sample_anno`
  argument.
  - `sample_summary`: sample copy number summary informations. See
    `data@cnv.summary` for details.
  - `cytoband_summary`: cytoband summary informations. See
    `data@cytoband.summary` for details.
  - `gene_summary`: gene summary informations. See
    `data@gene.summary` for details.
  - `summary`: A data frame of summary information. See `data@summary` for
    details.

```{r}
# Ensure the maftools package is installed and load the example GISTIC data
all.lesions <- system.file("extdata", "all_lesions.conf_99.txt", package = "maftools")
amp.genes <- system.file("extdata", "amp_genes.conf_99.txt", package = "maftools")
del.genes <- system.file("extdata", "del_genes.conf_99.txt", package = "maftools")
scores.gistic <- system.file("extdata", "scores.gistic", package = "maftools")
laml.gistic <- maftools::readGistic(
    gisticAllLesionsFile = all.lesions, gisticAmpGenesFile = amp.genes,
    gisticDelGenesFile = del.genes, gisticScoresFile = scores.gistic
)
ggoncoplot(laml.gistic) +
    scale_fill_brewer("CNV", palette = "Dark2", na.translate = FALSE) +
    theme_no_axes("x")
```
