# quad-layout {#sec-quad-layout}
`quad_layout()` arranges plots in the Quad-Side of a main plot.
```{r setup}
#| echo: false
#| output: false
library(ggalign)
set.seed(123)
small_mat <- matrix(rnorm(56), nrow = 7)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

```{r}
#| echo: false
ggside() +
    lapply(c("top", "left", "bottom", "right"), function(side) {
        c(
            list(quad_anno(side, initialize = TRUE)),
            lapply(seq_len(3L), function(i) {
                ggfree() +
                    ggtitle(sprintf("Plot %s", i)) +
                    theme(panel.background = element_rect(fill = scales::alpha("red", 0.5)))
            }),
            list(layout_annotation(theme = theme(
                plot.background = element_rect(color = "red", linewidth = unit(3, "cm"))
            )))
        )
    }) +
    quad_active() +
    ggtitle("Main Plot") +
    theme(panel.background = element_rect(fill = scales::alpha("red", 0.5))) +
    layout_title("QuadLayout") +
    layout_annotation(theme = theme(plot.title = element_text(size = 20)))
```

```{r setup}
#| code-fold: true
#| eval: true
```

## introduction
Depending on whether you want to align observations in the `horizontal` or `vertical`
direction, there are four main types of `quad_layout()`:

| Alignment of Observations     | horizontal | vertical | Data Format |
| ----------------------------- | ---------- | -------- | ----------- |
| `quad_free()`/`ggside()`      | ❌         | ❌       | data frame  |
| `quad_alignh()`               | ✅         | ❌       | matrix      |
| `quad_alignv()`               | ❌         | ✅       | matrix      |
| `quad_alignb()`/`ggheatmap()` | ✅         | ✅       | matrix      |

When the layout requires alignment of observations, we typically use a matrix,
regardless of whether alignment is needed in one or two directions.

 - For data frame, the input will be automatically converted using
 `fortify_data_frame()` if needed.
 - For matrix, the input will be automatically converted using
 `fortify_matrix()` if required.

## Annotations
Annotation is typically handled using a `stack_layout()`. Depending on whether
you want to align observations in the specified direction, different
`stack_layout()` are compatible (@sec-heatmap-layout-add-stack-layout). Below is
a table outlining the compatibility of various layout types for annotations:

| Annotations                   | left and right                   | top and bottom                   |
| ----------------------------- | -------------------------------- | -------------------------------- |
| `quad_free()`/`ggside()`      | `stack_freeh()`                  | `stack_freev()`                  |
| `quad_alignh()`               | `stack_freeh()`/`stack_alignh()` | `stack_freev()`                  |
| `quad_alignv()`               | `stack_freeh()`                  | `stack_freev()`/`stack_alignv()` |
| `quad_alignb()`/`ggheatmap()` | `stack_freeh()`/`stack_alignh()` | `stack_freev()`/`stack_alignv()` |

## `quad_alignb()`
`quad_alignb()` aligns observations in both horizontal and vertical directions. It
serves as the base version of `ggheatmap()`/`heatmap_layout()` and does not
automatically add default layers or mappings.

The underlying `ggplot` data of the main plot is the same with
`ggheatmap()`/`heatmap_layout()`, it is recommended to use `.y`, or `.discrete_y`
as the `y` mapping and use `.x` or `.discrete_x` as the `x` mapping in the
main plot.

```{r}
quad_alignb(small_mat, aes(.x, .y)) +
    geom_tile(aes(fill = value)) +
    scale_fill_viridis_c()
```

## `quad_alignh()`/`quad_alignv()`
`quad_alignh()` aligns observations horizontally, while `quad_alignv()` aligns
observations vertically.

The underlying `ggplot` data is the same with `ggheatmap()`/`heatmap_layout()`,
it is recommended to use `.y`, or `.discrete_y` as the `y` mapping in the main
plot for `quad_alignh()` but use `.x` or `.discrete_x` as the `x` mapping in
the main plot for `quad_alignv()`.

::: {.grid}

::: {.g-col-6}
```{r}
quad_alignh(small_mat) +
    geom_boxplot(aes(value, .discrete_y, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2") +
    layout_title("quad_alignh()")
```
:::

::: {.g-col-6}
```{r}
quad_alignv(small_mat) +
    geom_boxplot(aes(.discrete_x, value, fill = .column_names)) +
    scale_fill_brewer(palette = "Dark2") +
    layout_title("quad_alignv()")
```
:::

:::

As discussed in @sec-heatmap-layout-annotations, `quad_anno()` will always
attempt to initialize a `stack_layout()` with the same alignment as the current
direction. For top and bottom annotations in `quad_alignh()`, and left and right
annotations in `quad_alignv()`, `quad_anno()` will not initialize the annotation
due to inconsistent data types.

```{r}
quadh <- quad_alignh(small_mat) +
    anno_top()
quadv <- quad_alignv(small_mat) +
    anno_left()
```

Manual adding of a `stack_layout()` is required in such cases, you can set
`initialize = FALSE` to prevent the warning message.
```{r}
quadh <- quad_alignh(small_mat) +
    anno_top(initialize = FALSE)
quadv <- quad_alignv(small_mat) +
    anno_left(initialize = FALSE)
```

::: {.grid}

::: {.g-col-6}
```{r}
quadh +
    stack_freev(mpg) +
    # add a plot in the top annotation
    ggfree(mapping = aes(displ, hwy, colour = class)) +
    geom_point(aes(displ, hwy, colour = class)) +
    quad_active() +
    geom_boxplot(aes(value, .discrete_y, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2")+
    layout_title("quad_alignh()")
```
::: 

::: {.g-col-6}
```{r}
quadv +
    stack_freeh(data = mpg) +
    # add a plot in the top annotation
    ggfree(mapping = aes(displ, hwy, colour = class)) +
    geom_point(aes(displ, hwy, colour = class)) +
    quad_active() +
    geom_boxplot(aes(.discrete_x, value, fill = .column_names)) +
    scale_fill_brewer(palette = "Dark2") +
    layout_title("quad_alignv()")
```
::: 

::: 

Alternatively, you can set `initialize = TRUE`, which will initialize the
annotation stack layout with no data. In this case, you must provide `data` in
each plot within the annotation.

::: {.grid}

::: {.g-col-6}
```{r}
quad_alignh(small_mat) +
    geom_boxplot(aes(value, .discrete_y, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2") +
    anno_top(initialize = TRUE) +
    ggfree(data = mpg, aes(displ, hwy, colour = class)) +
    geom_point(aes(displ, hwy, colour = class))+
    layout_title("quad_alignh()")
```
::: 

::: {.g-col-6}
```{r}
quad_alignv(small_mat) +
    geom_boxplot(aes(.discrete_x, value, fill = .column_names)) +
    scale_fill_brewer(palette = "Dark2") +
    anno_left(initialize = TRUE) +
    ggfree(data = mpg, aes(displ, hwy, colour = class)) +
    geom_point(aes(displ, hwy, colour = class))+
    layout_title("quad_alignv()")
```
:::

:::

## `quad_free()`
`quad_free()` does not align observations and is functionally equivalent to the
[ggside](https://cran.r-project.org/package=ggside) package. For convenience,
`ggside()` is provided as an alias for quad_free(). This layout is particularly
useful for adding metadata or summary graphics along a continuous axis.

```{r}
ggside(mpg, aes(displ, hwy, colour = class)) +
    geom_point(size = 2) +
    # initialize top annotation
    anno_top(size = 0.3) +
    # add a plot in the top annotation
    ggfree() +
    geom_density(aes(displ, y = after_stat(density), colour = class), position = "stack") +
    # initialize right annotation
    anno_right(size = 0.3) +
    # add a plot in the right annotation
    ggfree() +
    geom_density(aes(x = after_stat(density), hwy, colour = class),
        position = "stack"
    ) &
    theme_bw()
```

`ggside()` allows facetting for the main plot, which should also be applied to
the annotations for proper alignment.
```{r}
i2 <- iris
i2$Species2 <- rep(c("A", "B"), 75)
ggside(i2, aes(Sepal.Width, Sepal.Length, color = Species)) +
    geom_point(size = 2) +
    facet_grid(Species ~ Species2) +
    anno_top(size = 0.3) +
    ggfree() +
    geom_density(aes(Sepal.Width, y = after_stat(density), colour = Species),
        position = "stack"
    ) +
    facet_grid(cols = vars(Species2)) +
    anno_right(size = 0.3) +
    ggfree() +
    geom_density(aes(x = after_stat(density), Sepal.Length, colour = Species),
        position = "stack"
    ) +
    facet_grid(rows = vars(Species)) &
    theme_bw()
```

If an annotation contains multiple plots, it can be tedious to add the same
element to each one individually. One way to simplify this is by creating an
external `stack_layout()` and adding the desired elements using the `&`
operator. Then, you can add this `stack_layout()` to the `quad_layout()`. In a
later chapter, I will introduce another operator that seamlessly combines with
the `+` operator, allowing you to add elements to multiple plots at once.


In the next chapter, we will explore even more advanced techniques for combining
multiple `quad_layout()`s. These methods will provide you with the tools to manage
more complex plot arrangements and make your visualizations even more flexible
and powerful.
