# stack layout {#sec-stack-layout}
`stack_layout()` arranges plots either horizontally or vertically. Based on
whether we want to align the discrete or continuous variables, there are two
types of stack layouts:

 - `stack_discrete()`: Align discrete variable along the stack.
 - `stack_continuous()`: Align continuous variable along the stack.

`stack_layout()` integrates the functionalities of `stack_discrete()` and
`stack_continuous()` into a single interface. The first argument for these three
functions is `direction` which should be a single string indicating the
direction of the stack layout, either `"h"`(`horizontal`) or `"v"`(`vertical`).

Several aliases are available for convenience:

 - `stack_discretev`: A special case of `stack_discrete` that sets `direction =
 "v"`.
 - `stack_discreteh`: A special case of `stack_discrete` that sets `direction =
 "h"`.
 - `stack_continuousv()`: A special case of `stack_continuous` that sets
 `direction = "v"`.
 - `stack_continuoush()`: A special case of `stack_continuous` that sets
 `direction = "h"`.

```{r setup}
#| echo: false
#| output: false
library(ggalign)
set.seed(123)
small_mat <- matrix(rnorm(56), nrow = 7)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

```{r}
#| echo: false
#| layout-ncol: 2
set.seed(123)
stack_discreteh(matrix(rnorm(56), nrow = 7)) +
    lapply(seq_len(3L), function(i) {
        ggalign() +
            ggtitle(sprintf("Plot %s", i)) +
            theme(panel.background = element_rect(fill = scales::alpha("red", 0.5)))
    }) +
    layout_title("stack_layout(direction = 'h')") +
    layout_theme(plot.title = element_text(size = 20))

stack_discretev(matrix(rnorm(56), nrow = 7)) +
    lapply(seq_len(3L), function(i) {
        ggalign() +
            ggtitle(sprintf("Plot %s", i)) +
            theme(panel.background = element_rect(fill = scales::alpha("red", 0.5)))
    }) +
    layout_title("stack_layout(direction = 'v')") +
    layout_theme(plot.title = element_text(size = 20))
```

```{r setup}
#| code-fold: true
#| eval: false
```

## Input data {#sec-stack-layout-input-data}
As discussed in @sec-input-data, when aligning discrete variables, we
typically use a matrix. For continuous axes, we can still use the long-formatted
data frame, which is the same as in ggplot2.

- For `stack_continuous()`, a data frame is required, and the input will be
automatically converted using `fortify_data_frame()` if needed. 
- For `stack_discrete()`, a matrix is required, and the input will be
automatically converted using `fortify_matrix()` if needed. 

By default, `fortify_data_frame()` will invoke the `ggplot2::fortify()` function
for conversion. Note, for matrix, it will be converted to a long-formatted data
frame which is different from the `ggplot2::fortify()`.

`stack_discrete()`/`stack_continuous()` will set up the layout, but no plot will
be drawn until you add a plot element:
```{r empty_layout}
#| eval: false
stack_discretev(small_mat) +                                                 # <1>
    layout_theme(plot.background = element_rect(color = "red"))              # <2>
# the same for `stack_continuous()`
```

  1. initialize a vertical stack layout.
  2. Add a plot background in for the entire layout.

```{r empty_layout}
#| echo: false
```

In this example, we use `layout_theme()` to insert a plot background in the
entire layout, it can also be used to control the theme of title,
subtitle, caption (`layout_title()`), guides, margins, panel.border, panel.spacing.

## Customizing the Layout {#sec-stack-layout-layout-customize}
When aligning discrete variables, the package offers a suite of `align_*`
functions designed to give you precise control over the ordering and groups.
Instead of detailing each `align_*` function individually, we will focus on the
general usage and how to combine them with `stack_discrete()`.

Here, we remain take `align_dendro()` as an example, it can reorder the
observations, split them into groups, and can add a plot for visualization.

When used for `stack_discreteh()`, the observations are aligned along the `y-axis`:
```{r}
stack_discreteh(small_mat) + # <1>
    align_dendro()           # <2>
```
  1. initialize a horizontal stack layout.
  2. reorder the observations based on the hierarchical clustering, add
  a dendrogram tree, and set the active plot to this plot.

When used for `stack_discretev()`, the observations are aligned along the `x-axis`:
```{r}
stack_discretev(small_mat) + # <1>
    align_dendro()        # <2>
```
  1. initialize a vertical stack layout.
  2. reorder the observations based on the hierarchical clustering, add
  a dendrogram tree, and set the active plot to this plot.

When `align_dendro()` is added to the layout, it performs following actions:

  1. reorder the observations.
  2. set the active plot to the dendrogram.

The active plot refers to the plot that subsequent ggplot2 components will
target. In this case, the active plot is the dendrogram, and any new layers
added will be applied to it. For instance, we can add additional layers to
visualize the dendrogram's structure or data. The default data underlying the
`ggplot` object of `align_dendro()` consists of the dendrogram node data. It is
also possible to use the dendrogram's edge data for customization, which I will
introduce in @sec-align-dendro.

```{r}
stack_discreteh(small_mat) + # <1>
    align_dendro() +         # <2>
    geom_point()             # <3>
```
  1. initialize a horizontal stack layout.
  2. reorder the observations based on the hierarchical clustering, add
  a dendrogram tree, and set the active plot to this plot.
  3. add a point layer to the dendrogram

The `active` argument controls whether a plot should be set as the active plot. It
accepts an `active()` object with the `use` argument to specify if the plot should
be active when added. 
```{r}
#| error: true
stack_discreteh(small_mat) +                             # <1>
    align_dendro(active = active(use = FALSE)) +      # <2>
    geom_point()                                      # <3>
```
  1. initialize a horizontal stack layout.
  2. reorder the observations based on the hierarchical clustering, add
  a dendrogram tree, but don't set the active plot to this plot.
  3. try to add a point layer to the dendrogram, should fail due to no active plot

Usually, you don't need to set this manually, as the active context is
automatically applied only for functions that add plot areas. You can inspect
whether a `align_*` function will add a plot by print it:
```{r}
align_dendro()
```

You might find it confusing that we mentioned `align_dendro()` will split
observations into groups, while the print output shows `split = "no"`. This
happens because we haven't specified the `k`/`h` argument in `align_dendro()`.
```{r}
align_dendro(k = 3L)
```

You don't need to explicitly provide `data` to `align_dendro()`. By default, it
inherits data from the layout. However, you can always provide another data
source, but note that this package uses the concept of `number of observations`
(`NROW()`). When aligning the observations, you must ensure the number of
observations is consistent across all plots.
```{r}
#| error: true
set.seed(123)
stack_discreteh(small_mat) +                                     # <1>
    align_dendro(data = matrix(rnorm(56), nrow = 8)) +        # <2>
    theme(axis.text.y = element_text())                       # <3>
```
  1. initialize a horizontal stack layout.
  2. reorder the observations based on hierarchical clustering, add a dendrogram
  tree, and set the active plot to this one, using self-provided data. This
  should fail because the number of observations is inconsistent.
  3. try to add y-axis text to the dendrogram.

```{r}
set.seed(123)
stack_discreteh(small_mat) +                                # <1>
    align_dendro(data = matrix(rnorm(70), nrow = 7)) +   # <2>
    theme(axis.text.y = element_text())                  # <3>
```
  1. initialize a horizontal stack layout.
  2. reorder the observations based on the hierarchical clustering, add
  a dendrogram tree, and set the active plot to this plot, using self-provided data
  3. add y-axis text to the dendrogram.

Alternatively, you can provide a function (or purrr-lambda) that will be applied
to the layout's matrix. For layouts that align observations, a matrix is always
required, so the data input must be in matrix form.
```{r}
set.seed(123)
stack_discreteh(small_mat) +                             # <1>
    align_dendro(data = ~ .x[sample(nrow(.x)), ]) +   # <2>
    theme(axis.text.y = element_text())               # <3>
```
  1. initialize a horizontal stack layout.
  2. reorder the observations based on the hierarchical clustering, add a
  dendrogram tree, and set the active plot to this plot, using self-provided
  data function
  3. add y-axis text to the dendrogram.

Without adding another plot, it's difficult to appreciate the benefits. Let's
now explore how to incorporate a plot.

## Initializing Plot {#sec-stack-layout-plot-initialize}
There are two primary functions for adding plots: 

  - `ggalign()`: Initialize a ggplot object and align the axes.
  - `ggfree()`: Initialize a ggplot object without aligning the axes.

Both functions initialize a `ggplot` object and, by default, set the `active` plot
when added to the layout.

```{r}
stack_discreteh(small_mat) +                                # <1>
    align_dendro() +                                        # <2>
    ggalign(data = rowSums) +                               # <3>
    geom_bar(aes(value, .discrete_y), stat = "identity") +  # <4>
    theme(axis.text.y = element_text())                     # <5>
```
  1. initialize a horizontal stack layout.
  2. reorder the observations based on the hierarchical clustering, add
  a dendrogram tree, and set the active plot to this plot.
  3. initialize a ggplot object, and set the active plot to this plot, using
  self-provided data function
  4. add a bar to the plot
  5. add y-axis text

You can build the plot separately and then add it to the layout:
```{r}
my_bar <- ggalign(data = rowSums) +                           # <1>
    geom_bar(aes(value, .discrete_y), stat = "identity") +    # <1>
    theme(axis.text.y = element_text())                       # <1>
stack_discreteh(small_mat) +                                  # <2>
    align_dendro() +                                          # <3>
    my_bar                                                    # <4>
```
  1. Create the bar plot.
  2. initialize a horizontal stack layout.
  3. reorder the observations based on the hierarchical clustering, add
  a dendrogram tree, and set the active plot to this plot.
  4. Add another bar plot to the layout.

The `active` argument can also control the place of the plot area to be added. It
accepts an `active()` object with the `order` argument to specify the order of
the plot area.
```{r}
stack_discreteh(small_mat) +
    align_dendro() +
    ggalign(data = rowSums, active = active(order = 1)) +
    geom_bar(aes(value, .discrete_y), stat = "identity") +
    theme(axis.text.y = element_text()) 
```

You can also stack plots vertically using `stack_discretev()`:
```{r}
stack_discretev(small_mat) + 
    align_dendro() +
    ggalign(data = rowSums) +
    geom_bar(aes(.discrete_x, value), stat = "identity") +
    theme(axis.text.y = element_text()) 
```

For `stack_continuous()`, no special action is taken except ensuring consistent
axis limits across all plots. You can manually control limits using
`continuous_limits()`:

::: {.grid}

::: {.g-col-6}
```{r}
stack_continuousv(mpg) +
    ggalign(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2) +
    ggalign(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2") &
    theme(
        panel.background = element_rect(fill = NA, color = "black"),
        axis.text.x = element_text()
    )
```
:::

::: {.g-col-6}
```{r}
stack_continuousv(mpg) +
    continuous_limits(c(3, 6)) +
    ggalign(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2) +
    ggalign(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2") &
    theme(
        panel.background = element_rect(fill = NA, color = "black"),
        axis.text.x = element_text()
    )
```
:::

:::

The `&` operator applies the added element to all plots in the layout, similar to
its functionality in the `patchwork` package.

`ggfree()` focuses on layout integration without enforcing strict axis alignment.

::: {.grid}

::: {.g-col-6}
```{r}
stack_discretev() +
    ggfree(mpg, aes(displ, hwy, colour = class)) +
    geom_point(size = 2) +
    ggfree(mpg, aes(displ, hwy, colour = class)) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2")  &
    theme(
        panel.background = element_rect(fill = NA, color = "black"),
        axis.text.x = element_text()
    )
```
:::

::: {.g-col-6}
```{r}
stack_continuousv(mpg) +
    continuous_limits(c(3, 6)) +
    ggfree(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2) +
    ggfree(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2") &
    theme(
        panel.background = element_rect(fill = NA, color = "black"),
        axis.text.x = element_text()
    )
```
:::

:::

By default, `ggfree()` will also inherit data from the layout and call
`fortify_data_frame()` to convert the data to a data frame. So, note that if the
layout data is a matrix, it will be converted into a long-formatted data frame. 

## Plot Size {#sec-stack-layout-plot-size}
Both `ggalign()` and `ggfree()` functions have a `size` argument to control the
relative `width` (for horizontal stack layout) or `height` (for vertical stack
layout) of the plot's panel area.
```{r}
stack_continuousv(mpg) +
    ggfree(mapping = aes(displ, hwy, colour = class), size = 2) +
    geom_point(size = 2) +
    ggfree(mapping = aes(displ, hwy, colour = class), size = 1) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2") &
    theme_bw()
```

Alternatively, you can define an absolute size by using a `unit()` object:
```{r}
stack_continuousv(mpg) +
    ggfree(mapping = aes(displ, hwy, colour = class), size = unit(1, "cm")) +
    geom_point(size = 2) +
    ggfree(mapping = aes(displ, hwy, colour = class)) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2") &
    theme_bw()
```

## active plot {#sec-stack-layout-active-plot}
As mentioned earlier, the active plot refers to the plot that subsequent ggplot2
components will target. The package provide two functions to work with active
plot for `stack_layout`.

  - `stack_switch()`: switch the active context
  - `stack_active()`: An alias for `stack_switch()`, which sets `what = NULL`

The `stack_switch()` function accepts the `what` argument, which can either be
the index of the plot added (based on its adding order) or the plot name
specified via the `active()` object using the `name` argument. 

Note that the `what` argument must be explicitly named, as it is placed second
in the function signature. This is because, in most cases, we don't need to
switch the active plot, manually—adjusting the order of plot additions typically
suffices.

::: {.grid}

::: {.g-col-6}
```{r}
stack_alignh(small_mat) +
    align_dendro() +
    ggalign(data = rowSums) +
    geom_bar(aes(value, .discrete_y), stat = "identity") +
    stack_switch(what = 1) +
    geom_point() +
    theme(axis.text.y = element_text()) +
    layout_title(title = "switch by integer")
```
:::

::: {.g-col-6}
```{r}
stack_alignh(small_mat) +
    align_dendro(active = active(name = "tree")) +
    ggalign(data = rowSums) +
    geom_bar(aes(value, .discrete_y), stat = "identity") +
    stack_switch(what = "tree") +
    geom_point() +
    theme(axis.text.y = element_text()) +
    layout_title(title = "switch by string")
```
:::

:::

In the example, we use `layout_title()` to insert a title for the entire layout.
Alternatively, you can add a title to a single plot with `ggtitle()`.

By setting `what = NULL` (or alias `stack_active()`), we remove the active plot.
This is particularly useful when the active plot is a nested `Layout` object, as
any additions would otherwise be directed to that nested `Layout`. By removing
the active plot, you can continue adding components to the `StackLayout`. We'll
introduce the nested layout of `StackLayout` in @sec-multiple-quad-layouts.

In the next chapter, we will dive into the `HeatmapLayout`, which can take the
`StackLayout` as input. Heatmap layouts offer additional features for aligning
observations in both directions. Let's move ahead and explore how heatmaps can
be seamlessly integrated into your layout workflows.
